Software architectures for workflow integration in science gateways

### 2016-04-21 paper.tex
Added a summary table in section 2

\usepackage[x11names,dvipsnames,table]{xcolor} %for use in color links
\usepackage[subrefformat=parens,labelformat=parens]{subfig}
\usepackage{svg}
\usepackage{url}
\usepackage{flushend}
\usepackage{xspace}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{multirow}

@@

In the last decade, the e-Science workflow community has developed
workflow systems to help developers access distributed infrastructures
such as clusters, grids, clouds and web services. These efforts
resulted in tools such as Askalon~\cite{fahringer2005askalon},
Hyperflow~\cite{balis2016hyperflow}, MOTEUR~\cite{GLAT-08i},
Pegasus~\cite{deelman2005pegasus,Deelman201517},
Swift~\cite{zhao2007swift}, Taverna~\cite{oinn2004taverna},
Triana~\cite{taylor2007triana}, VisTrails~\cite{callahan2006managing},
WS-PGRADE~\cite{Kacsuk2012}, etc. Such
workflow engines usually describe applications in a specific language
that offers parallel data and control flow constructs, visual edition
features, links with domain-specific tool repositories, provenance
recording, among other features.
% Descriptions and experiments
%conducted with e-Science workflow engines were published in journals
%such as Future Generation Computer Systems and conference venues such
%as WORKS.

@@

The architectures are diagramed in Figure~\ref{fig:architectures}
using the graphical notations shown in
Figure~\ref{fig:notations}. Architectures are described from their
main software \emph{components} and \emph{interactions}. Software
components include science gateway and infrastructure in all
architectures while workflow service, workflow pool and agent are
involved in some architectures. The workflow engine itself is
represented by a specific symbol with dotted lines. Software
interactions among these components are represented with
arrows. \emph{Abstract} interactions are specific types of
interactions that may be implemented by various different software
interactions. They are represented with dotted arrows and labeled with
 \texttt{*}. An architecture that has an abstract interaction is an
abstract architecture. Red color is used to represent the components
and interactions that are specific to workflow execution, i.e. that
are not involved in any other process. Tasks refer to computing tasks
that are created by the workflow engine and executed on the
infrastructure. Data represents any type of file, or database that is
involved in the workflow execution and stored on the infrastructure
while the workflow executes. It does not cover workflow parameters
such as strings, numbers, etc. The remainder of this section describes
the different types of interactions involved in the architectures, and
details each architecture.

@@

See Figure~\subref*{archi:tight}. The workflow engine is tightly
integrated with the science gateway, which means that it is deployed
on the same machine and potentially shares code, libraries and other
software components with the science gateway. For instance, the
workflow engine might be a portlet in a Liferay portal or a controller
in a Ruby on Rails application. The workflow engine and the science
gateway usually share a database where application, users and other
resources are stored. In this model, task and data control are both
initiated from the science gateway. Interactions \texttt{b} and
\texttt{c$_2$} are initiated from the workflow engine while
\texttt{c$_1$} comes from other parts of the science gateway, for
instance a data management interface. As in any other model, the
installation of new workflows in the science gateway (\texttt{a$_2$})
and infrastructure (\texttt{a$_1$}) is done by an administrator, for
security reasons. This is the model adopted in the Catania Science
Gateway Framework~\cite{Ardizzone2012} (see specific documentation on
workflows\footnote{\url{http://bit.ly/1oQrzvQ}}),
CIPRES~\cite{miller2010creating} and LONI Pipeline
Environment~\cite{dinov2009efficient}.
\todo{Add description of a real system.}

@@

This architecture is largely adopted, in systems such as Apache
Airvata~\cite{marru2011apache}, Vine
Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual
Imaging Platform~\cite{GLAT-13}, the WS-PGRADE/gUSE
framework~\cite{Kacsuk2012} and the numerous science gateway instances
that rely on
WS-PGRADE/gUSE~\cite{kacsuk2014science}. Figure~\ref{fig:vip-architecture}
shows the architecture of the Virtual Imaging Platform, where the
MOTEUR workflow engine~\cite{GLAT-08i} is invoked as a service at step
\texttt{3} (interaction \texttt{d} on
Figure~\ref{fig:architectures}). Step \texttt{2} on
Figure~\ref{fig:vip-architecture} corresponds to interaction
\texttt{c$_1$}, and step \texttt{4} maps to interactions \texttt{b}
and \texttt{c$_2$} (in VIP, workflow data transfers are performed by
the tasks and embedded in their descriptions).

@@

Nested workflows have long been available in workflow engines, for
instance in the Taverna workbench~\cite{oinn2004taverna}. They are
also used implicitly in several platforms where workflow engines are
wrapped in workflow tasks as any other command-line tool. Nested
workflows were notably used by the SHIWA Science Gateway to implement
so-called Coarse-Grained workflow
interoperability~\cite{terstyanszky2014enabling}, i.e. to integrate
various workflow engines in a consistent
platform. Figure~\ref{fig:shiwa-architecture} shows the architecture
used in the SHIWA Science Gateway for nested workflow execution with
service invocation as represented in Figure~\subref*{archi:nested}. The
parent workflow engine is \texttt{WS-PGRADE}, invoked as a service in
the Science Gateway (step \texttt{1} on
Figure~\ref{fig:shiwa-architecture}, interaction \texttt{d} on
Figure~\subref*{archi:nested}). Ten different child engines can be used by
nested workflows, invoked through the \texttt{Submission service}
(step \texttt{2}, interaction \texttt{d}). Each of these engines can
submit jobs and transfer data to a distributed computing
infrastructure (\texttt{DCI}, step \texttt{3}, interactions \texttt{b}
and \texttt{c$_2$}). Data interactions (\texttt{c}) and application
porting ones (\texttt{a}) are not represented on
Figure~\ref{fig:shiwa-architecture}.

@@

See Figure~\subref*{archi:import}. This is an abstract model that we
instantiate with the service invocation architecture for
consistency. Workflows are integrated in the science gateway through
format conversion from a native format to the science gateway
format. This was implemented in the SHIWA Science Gateway through
the IWIR language that provided a common language for portability
across grid workflow
systems~\cite{plankensteiner-prodan-etal:2013}.

\todo{workflow import is an offline process.}
\todo{Add description of a real system}

@@

The architectures described in Section~\ref{sec:architectures} are
evaluated in Table~\ref{table:evaluation}. We use 5 main criteria to
evaluate the architectures: integration effort, robustness of workflow
execution, extensibility, scalability and other features. Criteria
break down to specific metrics where \emph{low value indicates good
  performance}. For each criterion, a total score is computed by
summing up the individual metrics. We ensure that the different
metrics in a criterion measure similar entities so that summation
makes sense. The
criteria and metrics are explained hereafter.

@@

\emph{Other features} include:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Meta-workflow (\texttt{O$_1$}): ability to describe
  meta-workflows from existing workflows.  \todo{More details about
    why this is important?} \texttt{O$_1$}=0 when the feature is
  available, \texttt{O$_1$}=1 otherwise.
\item Fine-grained debugging (\texttt{O$_2$}): availability of
  fine-grained debugging information about workflow tasks. When
  such information is available, workflow is seen as a white
  box. Fine-grained information about workflow tasks is required to
  properly troubleshoot workflow executions.  \texttt{O$_2$}=0 when
  such information is easy to access, \texttt{O$_2$}=1
  otherwise. \todo{More details about why this is important?}
\end{itemize}
\texttt{O$_1$} and \texttt{O$_2$} are summed to obtain a total number of missing other
features in the architecture.  \todo{Maybe we could revise the
  categories to avoid such a vague ``other features''.}

@@

\paragraph{Other features} Meta-workflows are not supported by default
(\texttt{O$_1$}=1).  Debugging is not an issue since the science
gateway can retrieve any information from the workflow engine directly
(\texttt{O$_2$}=0).

@@

\paragraph{Other features} Meta-workflow are not supported by default
(\texttt{O$_1$}=1). Fine-grained debugging information is usually easy
to obtain since the workflow service provides direct access to the
engine (\texttt{O$_2$}=0).

@@

\paragraph{Other features} Meta-workflows are not available
(\texttt{O$_1$}=1).  Obtaining fine-grained information about workflow
tasks is not straightforward since the science gateway has no
knowledge about the workflow topology, and the workflow engine is
integrated as a task (\texttt{O$_2$}=1).

@@

\paragraph{Scalability} New engine instances only require new agents,
which is easily automated (\texttt{S$_1$=0}) and by design very
suitable for elastic computing. For instance, auto-scaling rules can
be implemented to start new agents when the workload in the science
gateway exceeds a certain threshold \todo{ref needed}. Distributed
engines are not available by default (\texttt{S$_2$=1}) and task
scheduling is as complex as in any other architecture
(\texttt{S$_3$=0}).

\paragraph{Other features} Meta-workflows are not available by default
(\texttt{O$_1$=1}).  Accessing debugging information is not likely to
be an issue since the workflow pool could implement specific functions
for that (\texttt{O$_2$=0}).

@@

\paragraph{Other features} 
Meta-workflows are possible, which is one of the main interest of this
architecture (\texttt{O$_1$=0}).  Debugging is difficult because the
science gateway cannot directly access fine-grained information in the
sub-workflow (\texttt{O$_2$=1}).

@@

\paragraph{Other features} Meta-workflows are available after import,
by connecting workflows in the language used in the science gateway
(\texttt{O$_1$=0}). Debugging information is accessible as in the
service invocation architecture (\texttt{O$_2$=0}).

@@

\multicolumn{7}{c}{\cellcolor[HTML]{EEEEEE}\textbf{Other features}}\\

@@

  \textbf{Total} (missing other features) & \cellcolor[HTML]{99DD99}\textbf{1}

@@

Differences in other features should not be neglected. Nested
workflows (with service invocation) and workflow import (with service
invocation) are the only architectures that support meta-workflows,
which may be interesting in some cases. The availability of
fine-grained debugging information may also be critical to efficient
user support.

Table~\ref{table:overall} provides an overall comparison between
architectures, based on the metrics in
Table~\ref{table:evaluation}. While Table~\ref{table:overall} provides
a global comparison between architectures, it should not be used
without the detailed metrics reported on Table~\ref{table:evaluation}
and discussed above. Overall, sub-tasking, and workflow import (with
service invocation) perform a bit better than tight integration,
service invocation and pool. Nested workflows stands a bit behind for
the reasons mentioned previously.
\begin{table*}
\centering
\begin{tabular}{rcccccc}
                                    & \textbf{Tight}
                                    & \textbf{Service}
                                    & \textbf{Sub-task}
                                    & \textbf{Pool}
                                    & \textbf{Nested}
                                    & \textbf{Import} \\
  Integration effort (normalized total) & 
                                    \cellcolor[HTML]{99FF99}{0.00}
                                    & \cellcolor[HTML]{99EF99}{0.22}
                                    & \cellcolor[HTML]{99FF99}{0.00}
                                    & \cellcolor[HTML]{99E099}{0.44}
                                    & \cellcolor[HTML]{99BB99}{1.00}
                                      & \cellcolor[HTML]{99E899}{0.33}\\                           
Robustness (normalized total) & 
                                \cellcolor[HTML]{99F599}{0.14}
                                    & \cellcolor[HTML]{99E199}{0.43}
                                    & \cellcolor[HTML]{99FF99}{0.00}
                                    & \cellcolor[HTML]{99CE99}{0.71}
                                    & \cellcolor[HTML]{99BB99}{1.00}
                                    & \cellcolor[HTML]{99E199}{0.43}\\
  Extensibility (normalized total)  & \cellcolor[HTML]{99E099}{0.44}
                                     & \cellcolor[HTML]{99E099}{0.44}
                                     & \cellcolor[HTML]{99FF99}{0.00}
                                     & \cellcolor[HTML]{99EF99}{0.22}
                                     & \cellcolor[HTML]{99BB99}{1.00}
                                     & \cellcolor[HTML]{99EF99}{0.22}\\
Scalability (normalized total)  & \cellcolor[HTML]{99BB99}{1.00}
                                     & \cellcolor[HTML]{99DD99}{0.50}
                                     & \cellcolor[HTML]{99DD99}{0.50}
                                     & \cellcolor[HTML]{99FF99}{0.00}
                                     & \cellcolor[HTML]{99DD99}{0.50}
                                     & \cellcolor[HTML]{99DD99}{0.50}\\
Other features (normalized total) & \cellcolor[HTML]{99DD99}{0.50}
                                     & \cellcolor[HTML]{99DD99}{0.50}
                                     & \cellcolor[HTML]{99BB99}{1.00}
                                     & \cellcolor[HTML]{99DD99}{0.50}
                                     & \cellcolor[HTML]{99DD99}{0.50}
                                     & \cellcolor[HTML]{99FF99}{0.00}\\
                                    & \cellcolor[HTML]{99EE99}\textbf{2.1}
                                    & \cellcolor[HTML]{99EE99}\textbf{2.1}
                                    & \cellcolor[HTML]{99FF99}\textbf{1.5}
                                    & \cellcolor[HTML]{99F499}\textbf{1.9}
                                    & \cellcolor[HTML]{99BB99}\textbf{4.0}
                                    & \cellcolor[HTML]{99FF99}\textbf{1.5}\\
\end{tabular}
\caption{Overall evaluation. Brighter colors and lower scores indicate better performance. Scores
  are obtained by summing the normalized total scores ($m'$ values) of
  each criterion in Table~\ref{table:evaluation} and the colors are obtained by normalizing these scores
  as done in Table~\ref{table:evaluation}. }
\label{table:overall}
\end{table*}

>>>>

\usepackage[x11names,dvipsnames,table]{xcolor} %for use in color links
\usepackage[subrefformat=parens,labelformat=parens]{subfig}
\usepackage{svg}
\usepackage{url}
\usepackage{flushend}
\usepackage{xspace}
\usepackage{enumitem}
\usepackage{hyperref}
\usepackage{multirow}
\usepackage{pbox}

@@

In the last decade, the e-Science workflow community has developed
workflow systems to help developers access distributed infrastructures
such as clusters, grids, clouds and web services. These efforts
resulted in tools such as Askalon~\cite{fahringer2005askalon},
Hyperflow~\cite{balis2016hyperflow}, MOTEUR~\cite{GLAT-08i},
Pegasus~\cite{deelman2005pegasus,Deelman201517},
Swift~\cite{zhao2007swift}, Taverna~\cite{oinn2004taverna},
Triana~\cite{taylor2007triana}, VisTrails~\cite{callahan2006managing},
WS-PGRADE~\cite{Kacsuk2012}, etc. Such workflow engines usually
describe applications in a specific language that offers parallel data
and control flow constructs, visual edition features, links with
domain-specific tool repositories, provenance recording, among other
features (see review in~\cite{deelman2009workflows}).
% Descriptions and experiments
%conducted with e-Science workflow engines were published in journals
%such as Future Generation Computer Systems and conference venues such
%as WORKS.

@@

The architectures are diagrammed in Figure~\ref{fig:architectures}
using the graphical notations shown in
Figure~\ref{fig:notations}. Architectures are described from their
main software \emph{components} and \emph{interactions}. Software
components include science gateway and infrastructure in all
architectures while workflow service, workflow pool and agent are
involved in some architectures. The workflow engine itself is
represented by a specific symbol with dotted lines. Software
interactions among these components are represented with arrows.
\emph{Abstract} interactions are specific types of interactions that
may be implemented by various different software interactions. They
are represented with dotted arrows and labeled with \texttt{*}. An
architecture that has an abstract interaction is an abstract
architecture.  Red color is used to represent the components and
interactions that are specific to workflow execution, i.e. that are
not involved in any other process. Tasks refer to computing tasks that
are created by the workflow engine and executed on the
infrastructure. Data represents any type of file, or database that is
involved in the workflow execution and stored on the infrastructure
while the workflow executes. It does not cover workflow parameters
such as strings, numbers, etc.

The remainder of this section describes the different types of
interactions involved in the architectures, details architectures and
illustrates them using examples of real
systems. Table~\ref{table:system-classification} summarizes the
classification of a few systems by architecture.

@@

See Figure~\subref*{archi:tight}. The workflow engine is tightly
integrated with the science gateway, which means that it is deployed
on the same machine and potentially shares code, libraries and other
software components with the science gateway. For instance, the
workflow engine might be a portlet in a Liferay portal or a controller
in a Ruby on Rails application. The workflow engine and the science
gateway usually share a database where application, users and other
resources are stored. In this model, task and data control are both
initiated from the science gateway. Interactions \texttt{b} and
\texttt{c$_2$} are initiated from the workflow engine while
\texttt{c$_1$} comes from other parts of the science gateway, for
instance a data management interface. As in any other model, the
installation of new workflows in the science gateway (\texttt{a$_2$})
and infrastructure (\texttt{a$_1$}) is done by an administrator, for
security reasons. This is the model adopted in LONI Pipeline
Environment~\cite{dinov2009efficient} and in the Catania Science
Gateway Framework~\cite{Ardizzone2012} which is used for instance in
the DECIDE gateway instance for
neuroimaging~\cite{ardizzone2012decide} (see specific documentation on
workflows\footnote{\url{http://bit.ly/1oQrzvQ}}). \todo{Add
  description of a real system.\note{Tristan}{The Catania Science
    Gateway Framework would be a good illustration here (workflow
    engine is a Java portlet), but I can't find a suitable figure
    about it.}}

%CIPRES~\cite{miller2010creating} (on which the Neuroscience gateway
%described in~\cite{sivagnanam2015early} is based) % \note{Tristan}{There is no workflow engine in there.}

@@

This architecture is largely adopted, in systems such as Apache
Airvata~\cite{marru2011apache}, Vine
Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual
Imaging Platform~\cite{GLAT-13}, the system
in~\cite{wu2010accelerating}, the WS-PGRADE/gUSE
framework~\cite{Kacsuk2012} and the numerous science gateway instances
that rely on WS-PGRADE/gUSE, for instance
e-BioInfra~\cite{shahand2015data}, MoSGrid~\cite{krüger2014mosgrid},
and many others~\cite{kacsuk2014science}. The
integration between the HubZero science gateway and Pegasus workflow
engine performed in~\cite{CPE:CPE3257} also use a service architecture
where the \texttt{d} interaction is implemented as a set of calls to
the Pegasus command-line tools (e.g. pegasus-status,
pegasus-statistics, etc). Figure~\ref{fig:vip-architecture} shows the
architecture of the Virtual Imaging Platform, where the MOTEUR
workflow engine~\cite{GLAT-08i} is invoked as a service at step
\texttt{3} (interaction \texttt{d} on
Figure~\ref{fig:architectures}). Step \texttt{2} on
Figure~\ref{fig:vip-architecture} corresponds to interaction
\texttt{c$_1$}, and step \texttt{4} maps to interactions \texttt{b}
and \texttt{c$_2$} (in VIP, workflow data transfers are performed by
the tasks and embedded in their descriptions).

@@

Nested workflows have long been available in workflow engines, for
instance in the Taverna workbench~\cite{oinn2004taverna}. They are
also used implicitly in several platforms where workflow engines are
wrapped in workflow tasks as any other command-line tool. Nested
workflows were notably used by the SHIWA Science Gateway to implement
so-called Coarse-Grained workflow
interoperability~\cite{terstyanszky2014enabling}, i.e. to integrate
various workflow engines in a consistent
platform. Figure~\ref{fig:shiwa-architecture} shows the architecture
used in the SHIWA Simulation Platform for nested workflow execution with
service invocation as represented in Figure~\subref*{archi:nested}. The
parent workflow engine is \texttt{WS-PGRADE}, invoked as a service in
the Science Gateway (step \texttt{1} on
Figure~\ref{fig:shiwa-architecture}, interaction \texttt{d} on
Figure~\subref*{archi:nested}). Ten different child engines can be used by
nested workflows, invoked through the \texttt{Submission service}
(step \texttt{2}, interaction \texttt{d}). Each of these engines can
submit jobs and transfer data to a distributed computing
infrastructure (\texttt{DCI}, step \texttt{3}, interactions \texttt{b}
and \texttt{c$_2$}). Data interactions (\texttt{c}) and application
porting ones (\texttt{a}) are not represented on
Figure~\ref{fig:shiwa-architecture}.

@@

See Figure~\subref*{archi:import}. This is an abstract model that we
instantiate with the service invocation architecture for
consistency. Workflows are integrated in the science gateway through
format conversion from a native format to the science gateway
format. Workflow import is usually an offline process which is not
involved in workflow execution. Workflow import was implemented in the SHIWA
Simulation Platform through the IWIR language that provided a common
language for portability across grid workflow
systems~\cite{plankensteiner-prodan-etal:2013}.

\todo{Add description of a real system}

\todo{Refer to ``From the desktop to the grid: scalable bioinformatics via workflow conversion Luis de la GarzaEmail author, Johannes Veit, Andras Szolek, Marc Röttig, Stephan Aiche, Sandra Gesing, Knut Reinert and Oliver Kohlbacher''}

\begin{table*}
\centering
\rowcolors{1}{white}{gray!25}
\begin{tabular}{ll}
  \textbf{Architecture} & \textbf{System} \\
  \hline
  Tight & \pbox{1.5\columnwidth}{
          Catania Science Gateway Framework~\cite{Ardizzone2012}, 
          DECIDE~\cite{ardizzone2012decide}$^{c,n}$, LONI Pipeline Environment$^n$~\cite{dinov2009efficient}
          }
  \\
  Service & \pbox{1.5\columnwidth}{
            Apache Airvata~\cite{marru2011apache}, e-BioInfra~\cite{shahand2015data}$^{w,n}$, HubZero with Pegasus~\cite{CPE:CPE3257}, MoSGrid~\cite{krüger2014mosgrid}$^w$, System in~\cite{wu2010accelerating}, Vine Toolkit~\cite{DBLP:journals/scpe/SzejnfeldDKKKKLPTWDNW10}, Virtual Imaging Platform~\cite{GLAT-13}$^n$, WS-PGRADE/gUSE framework~\cite{Kacsuk2012}, 
            Science gateways in ~\cite{kacsuk2014science}$^w$
            } \\
  Sub-task & CBRAIN and PSOM~\cite{GLAT-16}$^n$, CBRAIN and FSL$^n$\\
  Pool & SHIWA pool~\cite{ROGE-13}\\
  Nested & SHIWA Simulation Platform (Coarse-Grained Interoperability~\cite{terstyanszky2014enabling})$^w$\todo{others}\\
  Import & SHIWA Simulation Platform (Fine-Grained Interoperability)~\cite{plankensteiner-prodan-etal:2013}$^w$
\end{tabular}
\caption{Classification of science gateway based on the architecture used to integrate workflow engines. $^c$: based on the Catania Science Gateway Framework. $^w$: based on WS-PGRADE/gUSE. $^n$: used for neuroimaging.}
\label{table:system-classification}
\end{table*}

@@

The architectures described in Section~\ref{sec:architectures} are
evaluated in Table~\ref{table:evaluation}. We use 5 main criteria to
evaluate the architectures: integration effort, robustness of workflow
execution, extensibility, scalability and other specific features. Criteria
break down to specific metrics where \emph{low value indicates good
  performance}. For each criterion, a total score is computed by
summing up the individual metrics. We ensure that the different
metrics in a criterion measure similar entities so that summation
makes sense. The
criteria and metrics are explained hereafter.

@@

\emph{Specific features} include:
\begin{itemize}[leftmargin=0cm,itemindent=0.35cm,itemsep=0cm]
\item Meta-workflow (\texttt{O$_1$}): ability to describe
  meta-workflows from existing workflows.  \todo{More details about
    why this is important?} \texttt{O$_1$}=0 when the feature is
  available, \texttt{O$_1$}=1 otherwise.
\item Fine-grained debugging (\texttt{O$_2$}): availability of
  fine-grained debugging information about workflow tasks. When
  such information is available, workflow is seen as a white
  box. Fine-grained information about workflow tasks is required to
  properly troubleshoot workflow executions.  \texttt{O$_2$}=0 when
  such information is easy to access, \texttt{O$_2$}=1
  otherwise. \todo{More details about why this is important?}
\end{itemize}
\texttt{O$_1$} and \texttt{O$_2$} are summed to obtain a total number of missing specific
features in the architecture.

@@

\paragraph{Specific features} Meta-workflows are not supported by default
(\texttt{O$_1$}=1).  Debugging is not an issue since the science
gateway can retrieve any information from the workflow engine directly
(\texttt{O$_2$}=0).

@@

\paragraph{Specific features} Meta-workflow are not supported by default
(\texttt{O$_1$}=1). Fine-grained debugging information is usually easy
to obtain since the workflow service provides direct access to the
engine (\texttt{O$_2$}=0).

@@

\paragraph{Specific features} Meta-workflows are not available
(\texttt{O$_1$}=1).  Obtaining fine-grained information about workflow
tasks is not straightforward since the science gateway has no
knowledge about the workflow topology, and the workflow engine is
integrated as a task (\texttt{O$_2$}=1).

@@

\paragraph{Scalability} New engine instances only require new agents,
which is easily automated (\texttt{S$_1$=0}) and by design very
suitable for elastic computing. For instance, auto-scaling rules can
be implemented to start new agents when the workload in the science
gateway exceeds a certain threshold~\cite{lorido2012auto}. Distributed
engines are not available by default (\texttt{S$_2$=1}) and task
scheduling is as complex as in any other architecture
(\texttt{S$_3$=0}).

\paragraph{Specific features} Meta-workflows are not available by default
(\texttt{O$_1$=1}).  Accessing debugging information is not likely to
be an issue since the workflow pool could implement specific functions
for that (\texttt{O$_2$=0}).

@@

\paragraph{Specific features} 
Meta-workflows are possible, which is one of the main interest of this
architecture (\texttt{O$_1$=0}).  Debugging is difficult because the
science gateway cannot directly access fine-grained information in the
sub-workflow (\texttt{O$_2$=1}).

@@

\paragraph{Specific features} Meta-workflows are available after import,
by connecting workflows in the language used in the science gateway
(\texttt{O$_1$=0}). Debugging information is accessible as in the
service invocation architecture (\texttt{O$_2$=0}).

@@

\multicolumn{7}{c}{\cellcolor[HTML]{EEEEEE}\textbf{Specific features}}\\

@@

  \textbf{Total} (missing features) & \cellcolor[HTML]{99DD99}\textbf{1}

@@

Differences in other specific features should not be neglected. Nested
workflows (with service invocation) and workflow import (with service
invocation) are the only architectures that support meta-workflows,
which may be interesting in some cases. The availability of
fine-grained debugging information may also be critical to efficient
user support.

Table~\ref{table:overall} provides an overall comparison between
architectures, based on the metrics in
Table~\ref{table:evaluation}. While Table~\ref{table:overall} provides
a global comparison between architectures, it should not be used
without the detailed metrics reported on Table~\ref{table:evaluation}
and discussed above. Overall, sub-tasking, and workflow import (with
service invocation) perform a bit better than tight integration,
service invocation and pool. Nested workflows stands a bit behind for
the reasons mentioned previously.
\begin{table*}
\centering
\begin{tabular}{rcccccc}
                                    & \textbf{Tight}
                                    & \textbf{Service}
                                    & \textbf{Sub-task}
                                    & \textbf{Pool}
                                    & \textbf{Nested}
                                    & \textbf{Import} \\
  Integration effort (normalized total) & 
                                    \cellcolor[HTML]{99FF99}{0.00}
                                    & \cellcolor[HTML]{99EF99}{0.22}
                                    & \cellcolor[HTML]{99FF99}{0.00}
                                    & \cellcolor[HTML]{99E099}{0.44}
                                    & \cellcolor[HTML]{99BB99}{1.00}
                                      & \cellcolor[HTML]{99E899}{0.33}\\                           
Robustness (normalized total) & 
                                \cellcolor[HTML]{99F599}{0.14}
                                    & \cellcolor[HTML]{99E199}{0.43}
                                    & \cellcolor[HTML]{99FF99}{0.00}
                                    & \cellcolor[HTML]{99CE99}{0.71}
                                    & \cellcolor[HTML]{99BB99}{1.00}
                                    & \cellcolor[HTML]{99E199}{0.43}\\
  Extensibility (normalized total)  & \cellcolor[HTML]{99E099}{0.44}
                                     & \cellcolor[HTML]{99E099}{0.44}
                                     & \cellcolor[HTML]{99FF99}{0.00}
                                     & \cellcolor[HTML]{99EF99}{0.22}
                                     & \cellcolor[HTML]{99BB99}{1.00}
                                     & \cellcolor[HTML]{99EF99}{0.22}\\
Scalability (normalized total)  & \cellcolor[HTML]{99BB99}{1.00}
                                     & \cellcolor[HTML]{99DD99}{0.50}
                                     & \cellcolor[HTML]{99DD99}{0.50}
                                     & \cellcolor[HTML]{99FF99}{0.00}
                                     & \cellcolor[HTML]{99DD99}{0.50}
                                     & \cellcolor[HTML]{99DD99}{0.50}\\
Specific features (normalized total) & \cellcolor[HTML]{99DD99}{0.50}
                                     & \cellcolor[HTML]{99DD99}{0.50}
                                     & \cellcolor[HTML]{99BB99}{1.00}
                                     & \cellcolor[HTML]{99DD99}{0.50}
                                     & \cellcolor[HTML]{99DD99}{0.50}
                                     & \cellcolor[HTML]{99FF99}{0.00}\\
                                    & \cellcolor[HTML]{99EE99}\textbf{2.1}
                                    & \cellcolor[HTML]{99EE99}\textbf{2.1}
                                    & \cellcolor[HTML]{99FF99}\textbf{1.5}
                                    & \cellcolor[HTML]{99F499}\textbf{1.9}
                                    & \cellcolor[HTML]{99BB99}\textbf{4.0}
                                    & \cellcolor[HTML]{99FF99}\textbf{1.5}\\
\end{tabular}
\caption{Overall evaluation. Brighter colors and lower scores indicate better performance. Scores
  are obtained by summing the normalized total scores ($m'$ values) of
  each criterion in Table~\ref{table:evaluation} and the colors are obtained by normalizing these scores
  as done in Table~\ref{table:evaluation}. }
\label{table:overall}
\end{table*}
